---
title: "Multilevel Modeling Analysis"
author: "Jodie"
date: "2024-07-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Pre-processing

```{r cars}
library(readr)
library(pdp)
library(vip)
library(rpart.plot)
library(rpart)
library(caret)
library(tidyverse)
library(janitor)

# if there is an error, load data_org into your workspace by clicking on data_org copy.RData

sel_data<-health_data |>
  slice(-1) |>
  rename(
    Drug_Overdose_Deaths = drug_overdose_deaths_raw_value,
    mental_ratio = ratio_of_population_to_mental_health_providers,
    #uninsured=uninsured_raw_value,
    suicides = suicides_raw_value,
    Alcohol_Impaired_Driving_Deaths = alcohol_impaired_driving_deaths_raw_value,
    Unemployment_Rate = unemployment_raw_value,
    Income_Inequality = income_inequality_raw_value,
    College_Comp_Rate = some_college_raw_value,
    Adult_Smoking_Rate = adult_smoking_raw_value,
    Excessive_Drinking_Rate = excessive_drinking_raw_value,
    Median_House_Income = median_household_income_raw_value,
    Highschool_Comp_Rate = some_college_raw_value,
    Children_In_Poverty = children_in_poverty_raw_value,
    Single_Parent_House = children_in_single_parent_households_raw_value,
    Social_Associations = social_associations_raw_value,
    Pop_Under_18 = x_below_18_years_of_age_raw_value,
    Pop_Over_65 = x_65_and_older_raw_value,
    NH_Black = x_non_hispanic_black_raw_value,
    American_Or_Alaska_Native = x_american_indian_or_alaska_native_raw_value,
    Asian_Pop = x_asian_raw_value,
    Pacific_Islander = x_native_hawaiian_or_other_pacific_islander_raw_value,
    Hispanic_Pop = x_hispanic_raw_value,
    NH_White = x_non_hispanic_white_raw_value,
    English_Profic = x_not_proficient_in_english_raw_value,
    Female_Pop = x_female_raw_value,
    Rural_Pop = x_rural_raw_value
  ) 

selected_data <- sel_data|> # non-scaled
  select(
    name, state_abbreviation, x5_digit_fips_code, x5_digit_fips_code, Drug_Overdose_Deaths, Alcohol_Impaired_Driving_Deaths,
    Unemployment_Rate, Income_Inequality, Highschool_Comp_Rate, 
    Adult_Smoking_Rate, Excessive_Drinking_Rate, Median_House_Income,
    Highschool_Comp_Rate, Children_In_Poverty, Single_Parent_House,
    Social_Associations, Pop_Under_18, 
    Pop_Over_65, NH_Black, 
    American_Or_Alaska_Native, Hispanic_Pop, 
    Female_Pop, Rural_Pop, population_raw_value, mental_ratio, suicides, #uninsured
  ) |>
  mutate(x5_digit_fips_code=ifelse(nchar(x5_digit_fips_code) == 4, paste0("0", x5_digit_fips_code), x5_digit_fips_code)) |>
  mutate(across(c(Drug_Overdose_Deaths, Alcohol_Impaired_Driving_Deaths,
    Unemployment_Rate, Income_Inequality, Highschool_Comp_Rate, 
    Adult_Smoking_Rate, Excessive_Drinking_Rate, Median_House_Income,
    Highschool_Comp_Rate, Children_In_Poverty, Single_Parent_House,
    Social_Associations, Pop_Under_18, 
    Pop_Over_65, NH_Black, 
    American_Or_Alaska_Native, Hispanic_Pop, 
    Female_Pop, Rural_Pop, population_raw_value, mental_ratio, suicides), 
    as.numeric
  ))

# create a new variable that categorizes counties by HHS public health region 
selected_data <-selected_data |>
  mutate(region=case_when(
    state_abbreviation %in% c("CT", "ME", "MA", "NH", "RI", "VT") ~ "Region 1",
    state_abbreviation %in% c("NJ", "NY", "PR", "VI") ~ "Region 2",
    state_abbreviation %in% c("DE", "DC", "MD", "PA", "VA", "WV") ~ "Region 3",
    state_abbreviation %in% c("AL", "FL", "GA", "KY", "MS", "NC", "SC", "TN") ~ "Region 4",
    state_abbreviation %in% c("IL", "IN", "MI", "MN", "OH", "WI") ~ "Region 5",
    state_abbreviation %in% c("AR", "LA", "NM", "OK", "TX") ~ "Region 6",
    state_abbreviation %in% c("IA", "KS", "MO", "NE") ~ "Region 7",
    state_abbreviation %in% c("CO", "MT", "ND", "SD", "UT", "WY") ~ "Region 8",
    state_abbreviation %in% c("AZ", "CA", "HI", "NV", "AS", "MP", "FM", "GU", "MH", "PW") ~ "Region 9",
    state_abbreviation %in% c("AK", "ID", "OR", "WA") ~ "Region 10",
    TRUE ~ "Unknown"
  ))


library(tigris)
# us_counties df is strictly 5 numbers while selected_data has 4 digits for nums less than 10000

# us_counties <- counties(cb = TRUE, year=2022) # US COUNTIES, census_df have CA but selected_data doesn't have it (06037)
# census_df <- us_counties |>
#   select(x5_digit_fips_code=GEOID, land_area=ALAND)

# create population density variable
# new_sel <- merge(selected_data, census_df, by = "x5_digit_fips_code") |>
#   mutate(land_area_km = land_area / 1000000) |>
#   mutate(pop_den = population_raw_value / land_area_km) |>
#   mutate(urban_pop=1-Rural_Pop)

new_sel<-selected_data |>
  mutate(urban_pop=1-Rural_Pop)
  

quantiles <- quantile(new_sel$urban_pop, probs = c(0, 0.33, 0.66, 1))

# Display the quantiles
print(quantiles)

# create 2-level variable that represents if county is more rural or urban 
# create 3-level variable that represents if county is rural, suburban, or rural
new_df <- new_sel |>
  mutate(urban_level = ifelse(urban_pop==0, "rural", ifelse(urban_pop<0.52, "suburban", "urban"))) |>
  mutate(Urban_yes=ifelse(urban_pop>0.33, "urban", "rural")) # urban with just 2 levels

# high school--completed at least high school diploma
# some college--completed high school and pursued education after (not necessarily have to complete it)
# union--% that did not complete (1-high school)

# % of people under 18 in poverty (synonymous with income inequality)
library(caret)

# necessary to encode variables?

#dummyVars(~income_ine)

# generally recommended to scale your variables 

new_drugsdf <- new_df |>
  # mutate(income_ineq_level=ifelse(Income_Inequality<0.123, "low", "high")) |>
  mutate(urban_num=as.numeric(factor(urban_level, levels = c("rural", "suburban", "urban")))) |> 
  mutate(urban_num_twolev = as.numeric(factor(Urban_yes, levels = c("urban", "rural")))) |>
  select(x5_digit_fips_code, name, state_abbreviation, Drug_Overdose_Deaths, Unemployment_Rate, Income_Inequality, suicides, mental_ratio, Adult_Smoking_Rate, Median_House_Income, Highschool_Comp_Rate, Single_Parent_House, Social_Associations,
         NH_Black, American_Or_Alaska_Native, Hispanic_Pop, Female_Pop, region, urban_pop, urban_level,
         Urban_yes, urban_num, urban_num_twolev) |> # removed pop_den
  # select(-Alcohol_Impaired_Driving_Deaths, -Excessive_Drinking_Rate, 
  #        -geometry, -land_area_km, -population_raw_value, 
  #        -Rural_Pop, -land_area, -Highschool_Comp_Rate,
  #        -Pop_Under_18, -Pop_Over_65, -Children_In_Poverty, urban_level, region) |>
  mutate(across(c(Unemployment_Rate, Income_Inequality, Adult_Smoking_Rate, Median_House_Income, Highschool_Comp_Rate,
                  Social_Associations, NH_Black, American_Or_Alaska_Native, Hispanic_Pop, Female_Pop, suicides, mental_ratio), scale)) |>
  drop_na(Drug_Overdose_Deaths) 

new_drugsdf$urban_num_twolev<-as.factor(new_drugsdf$urban_num_twolev)
new_drugsdf$urban_num<-as.factor(new_drugsdf$urban_num)


# don't make this into a binary variable--urban_pop, urban_num, urban_num_twolev,
# urban_num_twolev--1 is urban, 2 rural

```

## Model Selection using AIC

You can also embed plots, for example:

```{r pressure, echo=FALSE}
new_drugsdf <- new_drugsdf |>
  select(Drug_Overdose_Deaths, Unemployment_Rate, Income_Inequality, Median_House_Income, Adult_Smoking_Rate, Highschool_Comp_Rate, suicides, mental_ratio, Social_Associations, Female_Pop, urban_num_twolev, urban_num, urban_level, region, state_abbreviation) |>
  drop_na()

library(lme4)
library(caret)
fixed_no_urban <-lm(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + Highschool_Comp_Rate + Social_Associations + Female_Pop + mental_ratio+ suicides, data=new_drugsdf)

fixed_urban <-lm(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + Highschool_Comp_Rate + Social_Associations + Female_Pop + urban_num_twolev+ mental_ratio+ suicides, data=new_drugsdf)

random_urban <-lmer(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + Highschool_Comp_Rate + Social_Associations + Female_Pop +  mental_ratio+ suicides +(1|urban_level), data=new_drugsdf)

random_state <-lmer(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + Highschool_Comp_Rate + Social_Associations + Female_Pop + urban_num_twolev + mental_ratio+ suicides+(1|state_abbreviation), data=new_drugsdf)

random_region <- lmer(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + Highschool_Comp_Rate + Social_Associations + Female_Pop + urban_num_twolev + mental_ratio+ suicides+ (1|region), data=new_drugsdf)

AIC(fixed_no_urban, fixed_urban, random_urban, random_region, random_state)

# arm::plot(fitted(random_state), residuals(random_state))
# 
# summary(random_state)
```

Using AIC score as a metric for model generalizability via leave one out cross validation, we can see that the model with state as random intercept has the lowest AIC.

## Checking model assumptions

```{r}
library(arm)
library(lme4)  # arm relies on lme4 for fitting mixed models
library(car)

plot(fitted(random_state), residuals(random_state))
#summary(random_state)

```

Fan shape in the fitted vs residuals plot, suggesting we need to apply a transformation. Try log transformation on the response variable

## Log-Transform Drug_overdose_deaths (response variable)

```{r}

log_random_state <-lmer(log(Drug_Overdose_Deaths)~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + Highschool_Comp_Rate + Social_Associations + Female_Pop + urban_num_twolev + mental_ratio+ suicides +(1|state_abbreviation), data=new_drugsdf)

plot(fitted(log_random_state), residuals(log_random_state))
# AIC(log_fixed_no_urban, log_fixed_urban, log_random_urban, log_random_state)

# log_random_s <-lmer(log(Drug_Overdose_Deaths)~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + Highschool_Comp_Rate + Social_Associations + Female_Pop + mental_ratio+ suicides +(1|state_abbreviation), data=selected_data)


```

## Plot (fixed) coefficient estimates 

```{r}
summary(log_random_state)
confint(log_random_state)

library(lmerTest)

model_random_state = summary(log_random_state)
coefficients <- model_random_state$coefficients[2:11, 1] 
conf_intervals <- confint(log_random_state, level = 0.95, oldNames=FALSE)

coef_df <- data.frame(
  Predictor = c('Unemployment Rate', 'Income Inequality',
                'Median Income', 'Smoking Rate', 'Highschool Comp Rate', 
                'Social Associations', '% Female', 'Urbanization', 'Mental Health Providers ratio', 'Suicide rate'),
  Estimate = coefficients,
  #StdError = coefficients[, "Std. Error"],
  ConfLow = conf_intervals[4:13, "2.5 %"],
  ConfHigh = conf_intervals[4:13, "97.5 %"]
)
# term = fct_reorder(term, estimate)


ggplot(coef_df, aes(x = Estimate, y = reorder(Predictor, Estimate), xmin = ConfLow, xmax = ConfHigh)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_errorbarh(height = 0.2, linewidth = 1.4, color = "red", show.legend = FALSE) +
  geom_point(size = 1.5, shape = 21, fill = "black", show.legend = FALSE) +
  labs(title = "Fixed Effects Estimates for Drug Overdose Deaths",
       x = "Coefficient",
       y = "Predictors") +
  labs(y = NULL) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16),      # Title text size
    axis.title.x = element_text(size = 14),                   # X-axis title text size
    axis.title.y = element_text(size = 14),                   # Y-axis title text size
    axis.text.x = element_text(size = 12),                    # X-axis text size
    axis.text.y = element_text(size = 12),                    # Y-axis text size
    legend.text = element_text(size = 12)                      # Legend text size
  )
  

abline(v = 0, col = "black", lwd = 2, lty = 2)
log_random_state
```

High suicide rate, % female, smoking rate, urbanization, income inequality is associated with high drug overdose. Low median income and low social associations is associated with low drug overdose.
urban has the most num of observations  in urban_num_two_lev

maybe use drug_overdose_deaths for each race 
Drug_Overdose_Deaths_asian=drug_overdose_deaths_aian,
    Drug_Overdose_Deaths_pacific=drug_overdose_deaths_asian_pacific_islander,
    Drug_Overdose_Deaths_black=drug_overdose_deaths_black
    
    
## Plot correlation matrix

```{r}
new_drugsdf_num <- new_drugsdf |>
  dplyr::select(!c(urban_num_twolev, region, state_abbreviation, urban_level, urban_num))

corr_mat<-cor(new_drugsdf_num)
# Plot the heat map
library(ggcorrplot)
ggcorrplot(corr_mat, method = "circle", type = "full", 
           lab = TRUE, lab_size = 3, 
           title = "Correlation Heat Map", 
           colors = c("blue", "white", "red"))

# library(ggplot2)
# ggplot(data=corr_mat, aes())

```
    
## Differences between rural and urban counties

```{r}
# two df--urban and rural counties
drugs_urban=new_drugsdf |>
  filter(urban_num_twolev==1)

drugs_rural=new_drugsdf |>
  filter(urban_num_twolev==2)

log_random_state_urban <-lmer(log(Drug_Overdose_Deaths)~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + Highschool_Comp_Rate + Social_Associations + Female_Pop + mental_ratio+ suicides +(1|state_abbreviation), data=drugs_urban)
log_state_urban <-lm(log(Drug_Overdose_Deaths)~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + Highschool_Comp_Rate + Social_Associations + Female_Pop + mental_ratio+ suicides, data=drugs_urban)

anova(log_random_state_urban, log_state_urban )
```
    
## Plot (random) coefficient estimates 

```{r}
library(lme4)
library(broom.mixed)
example(lmer)

# create get estimates effects of state 

# exclude fixed effects
# create a plot that visualizes the different intercepts for random state and CI

# conf.low and conf.high gives 95% confidence interval
tidy_model <- tidy(random_state, effects="ran_vals", conf.int = TRUE)

library(forcats)

# ADD THIS--reordering the levels
tidy_model$level <- fct_reorder(tidy_model$level, tidy_model$estimate)

ggplot(tidy_model, aes(y = estimate, x = level, ymin = conf.low, ymax = conf.high)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_errorbar(width = 0.2, linewidth = 1.4, color = "red", show.legend = FALSE) +
  geom_point(size = 1.5, shape = 21, fill = "black", show.legend = FALSE) +
  labs(title = "Random Effects Estimates for Drug Overdose Deaths",
       y = "Intercept Estimates",
       x = "State") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16),      # Title text size
    axis.title.x = element_text(size = 14),    # X-axis title text size
    axis.title.y = element_text(size = 14),    # Y-axis title text size
    axis.text.x = element_text(size = 6.7),     # X-axis text size
    axis.text.y = element_text(size = 12)      # Y-axis text size
  )
```

There is clear baseline variability in drug overdose deaths across states. Most notably, West Virginia has the the highest baseline drug overdose deaths

## Map of (random) coefficient estimates 

```{r}
# filtered_tidy <- tidy_model %>%
#   filter((conf.low > 0 & conf.high > 0) | (conf.low < 0 & conf.high < 0))
library(dplyr)
filtered_tidy_high <- tidy_model %>%
  filter((conf.low > 0 & conf.high > 0)) |>
  arrange(desc(level)) |>
  dplyr::select(level, estimate) # states that have baseline high intercepts in desc order

```


```{r}
# Install and load required packages
# install.packages("ggplot2")
# install.packages("maps")

library(ggplot2)
library(maps)

# Get the map data
us_map <- map_data("state")

# Create a data frame with states and values for coloring
set.seed(123)
states <- unique(us_map$region)
values <- runif(length(states), min = 1, max = 100) # do an if-else region
df_values <- data.frame(region = states, value = values)
df_values$level<- state.abb[match(tolower(df_values$region), tolower(state.name))]
df_values <- df_values %>%
  mutate(level = ifelse(is.na(level), "DC", level)) 

merged_df_values <- left_join(df_values, filtered_tidy_high, by="level")
# df_values[1]['abbreviation'] = 'DC'
# df_values <- df_values |>
#   filter(region == 'district of columbia') |>
#   select(abbreviation) |>
  

# Merge the map data with your data frame
us_map <- merge(us_map, merged_df_values, by = "region")

# Plot the map
ggplot(data = us_map, aes(x = long, y = lat, group = group, fill = estimate)) +
  geom_polygon(color = "black") +
  coord_fixed(1.3) +
  labs(fill = "Baseline Drug Overdoses") +
  theme_void() +
  theme(legend.position = "right")

```

States in midwest and east coast (region 3 and 4) have highest baseline drug overdoses--deaths not attributed to the predictors in the model. Thus, we should prioritize drug overdose prevention efforts on these states.  

plot drug overdoses across west virginia 